---
title: Preliminary Report
date: "`r Sys.Date()`"
output: 
  html_document:
          code_folding: hide
          code_download: true
          toc: true
          toc_float: true
  github_document:
          html_preview: false
          toc: true
          toc_depth: 3
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=T, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.align = "center",
  fig.path = "../Figures/markdown/",
  dev = c("png", "pdf"),
  message = FALSE,
  warning = FALSE,
  collapse = T
)

theme_matt = function(base_size = 18,
                      dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  ggpubr::theme_pubr(base_family="sans") %+replace% 
    theme(
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}

theme_matt_facets = function(base_size = 18,
                             dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  theme_bw(base_family="sans") %+replace% 
    theme(
      panel.grid = element_blank(),
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      strip.text.x = element_text(size = base_size),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}
```

## Temperature and Salinity in Lake Champlain

To provide context for this project, we begin by examining patterns in temperature and salinity in Lake Champlain.

### Acquiring environmental data 

The USGS maintains a continuous data record near Burlington Fishing Pier, where copepods were collected. This sensor records both temperature and specific conductivity.    

```{r}
#### Load the data ####
# Site code = 04294500 (this is the site in Lake Champlain near Burlington)
# Parameter code = 00010 (water temperature in degrees C)
# Parameter code = 00095 (specific conductivity at 25C)
# Service should be dv to get daily averaged values

siteNumber = "04294500"
parameterCd = c("00010", "00095")
startDate = ""
endDate = ""

url = constructNWISURL(siteNumbers = siteNumber, parameterCd = parameterCd, 
                       startDate = startDate, endDate = endDate, service = "dv")

env_data = importWaterML1(url, asDateTime = T, tz = "America/New_York") %>%  
  mutate("date" = as.Date(dateTime)) %>%  
  select(date, "temp" = X_00010_00003, "cond" = X_00095_00003) %>% 
  drop_na(temp, cond) %>% 
  mutate(doy = yday(date)) %>% 
  mutate(mgL = cond * 0.292) # State equation to convert continuous conductivity measurements to chloride concentrations: VT DEC 2019 - Watershed Management Division. Vermont Surface Water Assessment and Listing Methodology in accordance with USEPA Guidance. Montpelier www.watershedmanagement.vt.gov

```

Data for a total of `r dim(env_data)[1]` days is available, covering a period of time spanning from `r format(min(env_data$date), format="%B %d, %Y")` to `r format(max(env_data$date), format="%B %d, %Y")`.

### Seasonal patterns 

Shown below are the plots of the seasonal patterns in temperature and salinity. The temperature plot highlights that Lake Champlain is highly seasonal, with a \>20°C difference in temperature between Winter and Summer.

The salinity plot shows a different seasonal pattern, with slight increases in both winter and summer. These increases may be driven by different processes: increased salt input to the lake during Winter, and increased evaporative water loss during the Summer. Note that the highest salinity values are observed during the Winter, occuring as infrequent, short spikes in salinity.

```{r temp-sal-plots, fig.height=9, fig.width=6}

#Temperature climatology
temp_plot = ggplot(env_data, aes(x = doy, y = temp)) + 
  geom_point(alpha = 0.2, colour = "grey30") + 
  geom_smooth(colour = "black", linewidth = 2) + 
  labs(x = "", 
       y = "Temperature (°C)") + 
  theme_matt()

#Salinity climatology
sal_plot = ggplot(env_data, aes(x = doy, y = mgL)) + 
  geom_point(alpha = 0.2, colour = "grey30") + 
  geom_smooth(colour = "black", linewidth = 2) + 
  labs(x = "Day of Year", 
       y = "Salinity (mg/L)") + 
  theme_matt()

ggarrange(temp_plot, sal_plot, nrow = 2)
```

The correlation between temperature and salinity is weakly positive, but again we see that the highest salinity values occur when water temperatures are low. If we restrict the correlation to just days where temperature is below 10°C, we see that there is generally no relationship between salinity and temperature. The overall average salinity is included in this restricted plot in red. This suggests that Lake Champlain during the winter is largely characterized by stable salinities, with a small number of relatively extreme events.

```{r temp-salt-corrs, fig.width=6}

#Correlations
ggplot(env_data, aes(x = temp, y = mgL)) + 
  geom_point(alpha = 0.3, colour = "grey30") + 
  geom_smooth(colour = "black", linewidth = 2) + 
  labs(x = "Temperature (°C)", 
       y = "Salinity (mg/L)") + 
  theme_matt()

### Summer increase in conductivity driven by increasing evaporation / decreasing input? 
env_data %>% 
  filter(temp < 10) %>% 
  ggplot(aes(x = temp, y = mgL)) + 
  geom_hline(yintercept = mean(env_data$mgL, na.rm = T), linewidth = 2, colour = "sienna3") + 
  geom_point(alpha = 0.2, colour = "grey30") + 
  geom_smooth(colour = "black", linewidth = 2) + 
  labs(x = "Temperature (°C)", 
       y = "Salinity (mg/L)") + 
  theme_matt()
```

Shown below is the distribution of observed salinity values from the sensor, restricted to data from days were the temperature was less than or equal to 10°C. This distribution suggests that salinity levels are fairly low in the lake, with the distribution centered around 55 mg/L. 

```{r sal-hist, fig.width=6, include = F}
#The distribution of daily salinity values is shown below, along with vertical lines indicating the 50th, 75th, 90th, and 99th percentiles for winter days (water temperatures below 10°C). These quartile values would be strong candidates for **environmentally realistic** levels of salinity exposures.

winter_vals = env_data %>% 
  filter(temp <= 10)

sal_quants = quantile(winter_vals$mgL, probs = c(0.5, 0.75, 0.9, 0.99), na.rm = T)

ggplot(env_data, aes(x = mgL)) + 
  geom_histogram(binwidth = 2, 
                 fill = "grey30") + 
  #geom_vline(xintercept = sal_quants, colour = "steelblue3", linewidth = 1) + 
  labs(x = "Salinity (mg/L)") + 
  theme_matt()

```

## Preliminary Survival Experiments

To select salinity treatments for the full experiment we ran a multi-day survival test, with copepods exposed to a range of salinities. The first preliminary trial used concentrations based on the observed salinity distribution in the lake, and the EPA guidlines on acute and extended exposure thresholds. These values (0 mg/L - 1000 mg/L added to bottled spring water) span the observed salinities, and extend past the EPA thresholds.

Survival was checked approximately every 24 hours. After four days, upper thermal limits (as CTmax) were measured for five copepods each from the control and 1000 mg/L salinity treatment. 

### Survival Analysis   

Surprisingly, survival was high across all treatments in this trial, with only minimal mortality in the highest salinity treatments. 

```{r surv-trial, fig.width=6, fig.height=6}

surv_trial_1 = expand_surv(prelim_surv) %>% make_surv()
surv_obj_1 = Surv(surv_trial_1$hour, surv_trial_1$ind_surv)
surv_fit_1 = survfit2(Surv(hour, ind_surv) ~ treatment, data = surv_trial_1)

surv_trial_1  %>% 
  ungroup() %>% 
  select(-ID, -hour) %>% 
  group_by(treatment, initial) %>% 
  summarise(num_died = sum(ind_surv)) %>%  
  mutate(total_surv = initial - num_died,
         prop_surv = total_surv / initial) %>% 
  ggplot(aes(x = treatment, y = prop_surv)) + 
  geom_hline(yintercept = 0.5,
             colour = "grey", 
             linetype = "dashed") + 
  geom_point(size = 4) + 
  labs(x = "Salinity (mg/L)",
       y = "Proportion Surviving ") + 
  theme_matt()


#summary(surv_fit)

ggsurvplot(surv_fit_1, 
           conf.int=T, pval=F, risk.table=F, 
           conf.int.alpha = 0.1,
           size = 2,
           legend.title="Salt Treatment",
           palette=c("lightskyblue1", "dodgerblue","dodgerblue3", "dodgerblue4", "navy", "darkorchid4", "lightsalmon", "indianred2", "indianred4", "orangered3", "firebrick4"))
```

This first preliminary test suggests that *L. sicilis* has fairly high salinity tolerance. A second preliminary test was run with a much larger salinity range (spanning 1000 to 7000 mg salt/L). No acute mortality (<5 hours after the ramp up was completed) was observed, but within 24 hours there was mortality observed in some of the higher salt concentration treatments. Note that the salinity values where mortality occured correspond to >5 g added salt per L water. 

```{r trial-2-init-mort, fig.width=8, fig.height=6}

daily_surv_data = data.frame()

for(i in 3:dim(prelim_surv2)[2]){
  
  col_name = colnames(prelim_surv2)[i]
  hour = as.numeric(str_split_fixed(col_name, pattern = "_", n = 2)[2])
  
  day_data = prelim_surv2 %>%  
  select(treatment:{{col_name}}) %>% 
    expand_surv() %>% 
  make_surv() %>% 
  ungroup() %>% 
  group_by(treatment, initial) %>% 
  summarise(num_died = sum(ind_surv)) %>%  
  mutate(total_surv = initial - num_died,
         prop_surv = total_surv / initial,
         hour = hour)
  
  daily_surv_data = bind_rows(daily_surv_data, day_data)
}

ggplot(daily_surv_data, aes(x = treatment, y = prop_surv, colour = factor(hour))) + 
  geom_hline(yintercept = 0.5,
             colour = "grey", 
             linetype = "dashed") + 
  geom_point(size = 4) + 
  geom_smooth(method = "glm", 
    method.args = list(family = "binomial"), 
    se = FALSE) + 
  labs(x = "Salinity (mg/L)",
       y = "Proportion Surviving",
       colour = "Hour") + 
  theme_matt()
```

```{r include = F}
daily_surv_data = data.frame()

for(i in 3:dim(prelim_surv2)[2]){
  
  col_name = colnames(prelim_surv2)[i]
  hour = as.numeric(str_split_fixed(col_name, pattern = "_", n = 2)[2])
  
  day_data = prelim_surv2 %>%  
  select(treatment:{{col_name}}) %>% 
    expand_surv() %>% 
  make_surv() %>% 
  ungroup() %>% 
  group_by(treatment, initial) %>% 
  summarise(num_died = sum(ind_surv)) %>%  
  mutate(total_surv = initial - num_died,
         prop_surv = total_surv / initial,
         hour = hour)
  
  daily_surv_data = bind_rows(daily_surv_data, day_data)
}

prelim_surv2 %>%  
  expand_surv() %>% 
  mutate(ind_surv = -1*(ind_surv - 1)) %>% 
ggplot(aes(x = treatment, y = ind_surv, colour = factor(hour))) + 
  geom_hline(yintercept = 0.5,
             colour = "grey", 
             linetype = "dashed") + 
  geom_point(size = 4) + 
  geom_smooth(method = "glm", 
    method.args = list(family = "binomial"), 
    se = F) + 
  labs(x = "Salinity (mg/L)",
       y = "Proportion Surviving",
       colour = "Hour") + 
  theme_matt()
```


```{r surv-trial-2, fig.width=10, fig.height=6}
surv_trial_2 = expand_surv(prelim_surv2) %>% make_surv()
surv_obj_2 = Surv(surv_trial_2$hour, surv_trial_2$ind_surv)
surv_fit_2 = survfit2(Surv(hour, ind_surv) ~ treatment, data = surv_trial_2)

#summary(surv_fit_2)

ggsurvplot(surv_fit_2, 
           conf.int=T, pval=F, risk.table=F, 
           conf.int.alpha = 0.1,
           size = 2,
           legend.title="Salt Treatment")
```

```{r surv-model-2, fig.width=6, fig.height=6}
cox.model_2 = coxph(Surv(hour, ind_surv) ~ treatment, data = surv_trial_2)
print(cox.model_2)

#ggforest(cox.model_2, data = surv_trial_2)
```


### CTmax Analysis 

After several days of acclimation to the new salt levels, CTmax was measured for individual copepods. In trial 1, the control and maximum salt concentrations were used (0 and 1000 mg/L, respectively). One CTmax assay was performed (n = 5 per treatment), resulting in only a small comparison. Nonetheless, no large differences in CTmax were observed. 

```{r ctmax-trial, fig.width=6, fig.height=7}
ctmax_data2 = est_ctmax(ctmax_temp2, ctmax_time2)

ggplot(ctmax_data2, aes(x = treatment, y = ctmax)) + 
  geom_boxplot(width = 0.5) +
  geom_point(size = 4) + 
  labs(x = "Treatment", 
       y = "CTmax (°C)",
       title = "Trial 1") + 
  theme_matt()
```

```{r message = T}
t.test(data = ctmax_data2, ctmax~treatment)
```


