---
title: Effects of salinity acclimation on upper thermal limits of Lake Champlain diaptomid copepods
date: "`r Sys.Date()`"
output: 
  html_document:
          code_folding: hide
          code_download: true
          toc: true
          toc_float: true
  github_document:
          html_preview: false
          toc: true
          toc_depth: 3
---

```{r setup, include=T, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.align = "center",
  fig.path = "../Figures/report/",
  dev = c("png", "pdf"),
  message = FALSE,
  warning = FALSE,
  collapse = T
)

theme_matt = function(base_size = 18,
                      dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  ggpubr::theme_pubr(base_family="sans") %+replace% 
    theme(
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}

theme_matt_facets = function(base_size = 18,
                             dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  theme_bw(base_family="sans") %+replace% 
    theme(
      panel.grid = element_blank(),
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      strip.text.x = element_text(size = base_size),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}
```

## Survival Analyses

```{r}
ggplot(daily_prop_data, aes(x = treatment, y = prop_surv, colour = factor(exp_day))) + 
  facet_wrap(salt~.) + 
  geom_hline(yintercept = 0.5,
             colour = "grey", 
             linetype = "dashed") + 
  geom_point(size = 3) + 
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              se = FALSE,
              linewidth = 2) + 
  scale_colour_brewer(type = "seq", palette = 9) + 
  guides(colour = guide_legend(nrow = 1)) + 
  labs(x = "Salinity (mg/L)",
       y = "Proportion Surviving",
       colour = "Day") + 
  theme_matt_facets() + 
  theme(legend.position = "bottom")
```

```{r}
# surv_obj = Surv(surv_data$hour, surv_data$ind_surv)
# surv_fit = survfit2(Surv(hour, ind_surv) ~ treatment + salt, data = surv_data)
# 
# #summary(surv_fit_2)
# 
# ggsurvplot_facet(surv_fit, 
#                  data = surv_data,
#                  facet.by = "salt",
#                  conf.int=T, pval=F, risk.table=F, 
#                  conf.int.alpha = 0.1,
#                  size = 2,
#                  palette = "YlOrRd",
#                  legend.title="Salt Treatment")
```

```{r include = F}
cox.model = coxph(Surv(exp_day, ind_surv) ~ treatment + salt, data = surv_data)

cox.model

#ggforest(cox.model, data = surv_data)
```

### Skistodiaptomus oregonensis

```{r, fig.width=10, fig.height=6}

oreg_data %>% 
  mutate(initial = hour_0) %>% 
  pivot_longer(cols = c("hour_0", "hour_18", "hour_42", "hour_62", "hour_86"), 
               names_to = "hour", 
               values_to = "individuals") %>% 
  mutate(prop_surv = individuals / initial) %>% 
  ggplot(aes(x = treatment, y = prop_surv, colour = sex)) +
  facet_wrap(hour~.) + 
  geom_point(position = position_jitter(height = 0.01, width = 200)) + 
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              se = FALSE,
              linewidth = 2) + 
  scale_colour_manual(values = c("female" = "lightcoral", 
                                 "male" = "lightblue3")) + 
  labs(x = "Salinity (mg/L)", 
       y = "Proportion Surviving") + 
  theme_matt_facets()

```

## CTmax Data

```{r include = F}
ctmax_data %>%  
  mutate("ID" = paste(experiment_date, "- Exp.", experiment)) %>% 
  ggplot(aes(x = treatment, y = ctmax, fill = ID)) +
  facet_wrap(salt~.) +
  geom_boxplot(width = 0.5,
               position = position_dodge(width = 0.7)) +
  geom_point(size = 3,
             position = position_dodge(width = 0.7)) + 
  labs(x = "Treatment", 
       y = "CTmax (°C)") + 
  guides(fill = guide_legend(override.aes = list(shape = NA))) + 
  theme_matt() + 
  theme(legend.position = "right")
```

Shown below are the sample sizes for the various salt-treatment combinations. Fewer experiments were run with Instant Ocean, so the sample sizes are smaller for this salt.

```{r}
ctmax_filtered = ctmax_data %>% 
  filter(!(experiment == 1 & experiment_date == "1/20/24")) %>% 
  filter(!(experiment_date == "2/3/24")) %>% 
  mutate("ID" = paste(salt, treatment, sep = " - ")) %>% 
  mutate(group = case_when(
    treatment == "control" ~ "control",
    treatment == "salt" ~ salt,
    treatment == "salt" ~ salt
  ))

knitr::kable(ctmax_filtered %>% 
               group_by(salt, treatment) %>%  
               count())
```

A linear mixed effects model was used to examine variation in CTmax as a function of salt type (Instant Ocean and road salt) and treatment (control or salt-acclimated). We also examined the interaction between these factors to determine whether the effect of treatment depended on the type of salt used. Experiment date was included as a random effect to control for any differences between collections or experimental replicates.

```{r}
salt.model = lmer(data = ctmax_filtered,
                  ctmax ~ treatment * salt + (1|experiment_date))

# salt.model = lm(data = ctmax_filtered,
#                 ctmax ~ treatment * salt)

knitr::kable(car::Anova(salt.model, type = "III"))
```

The mixed effects model indicates that there was a significant interaction between salt type and treatment. We examined this interaction by calculating the marginal treatment means by salt type. Shown below, there was no significant difference between treatment groups in the Instant Ocean experiments, but there was a substantial difference between treatment groups in the road salt experiments.

```{r}
salt.means = emmeans::emmeans(salt.model,
                              pairwise ~ treatment | salt)

knitr::kable(salt.means$contrasts)
```

```{r plot-for-poster, fig.width = 8, fig.height=7}

treat_cols = c("salt" = "#95d5ce", "control" = "#028260")

ggplot(ctmax_filtered, aes(x = treatment, y = ctmax, fill = treatment)) +
  facet_wrap(salt~.) + 
  geom_boxplot(width = 0.5) +
  geom_point(size = 4, alpha = 0.5) + 
  scale_fill_manual(values = treat_cols) + 
  labs(x = "Treatment", 
       y = "CTmax (°C)") + 
  theme_matt_facets() + 
  theme(legend.position = "none",
        strip.background = element_rect(fill = "lightsalmon"),
        strip.text = element_text(margin = margin(0.7,0,0.7,0, "cm"),
                                  size = 20))

# ggplot(ctmax_filtered, aes(x = ctmax, fill = treatment)) + 
#   facet_wrap(salt~.) + 
#   geom_density(alpha = 0.5)
# 
# # Approach 1 - Simple t-test (comparison between two groups)
# t.test(data = filter(ctmax_filtered, salt == "RoadSalt"), ctmax ~ treatment)
```

```{r include = F}
ggplot(ctmax_filtered, aes(x = treatment, y = ctmax, fill = treatment)) +
  facet_wrap(salt~.) + 
  geom_violin(draw_quantiles = c(0.25,0.75)) + 
  geom_point(size = 4) + 
  labs(x = "Treatment", 
       y = "CTmax (°C)") + 
  theme_matt()
```

## Ion Specific Patterns Across Seasons

### Sample Sizes 

```{r}

species_surv_total = surv_2025_data %>% 
  group_by(species, treatment) %>% 
  count(name = "total")

species_surv_collections = surv_2025_data %>% 
  select(species, treatment, collection_date) %>% 
  distinct() %>% 
  group_by(species, treatment) %>% 
  count(name = "collections")

species_surv_mean = surv_2025_data %>% 
  group_by(species, treatment, collection_date) %>% 
  count() %>% 
  ungroup() %>% 
  group_by(species, treatment) %>% 
  summarise(mean_n = mean(n))

species_surv_concentrations = surv_2025_data %>% 
  select(species, treatment, collection_date, cl_conc) %>% 
  distinct() %>% 
  group_by(species, treatment, collection_date) %>% 
  count() %>% 
  ungroup() %>% 
  group_by(species, treatment) %>% 
  summarise(mean_n_conditions = mean(n))


species_surv_total %>% 
  inner_join(species_surv_collections, by = c("species", "treatment")) %>% 
  inner_join(species_surv_mean, by = c("species", "treatment")) %>% 
  inner_join(species_surv_concentrations, by = c("species", "treatment")) %>% 
  knitr::kable(digits = 2, caption = "Sample size information from the survivorship experiments. Shown here for each species and salt type are the total number of individuals monitored for survival, the number of collections (i.e. experimental replicates), and the average number of individuals per replicate, and the average number of conditions (chloride concentrations) employed during each replicate.")

```



```{r}

species_ctmax_total = acclim_data %>% 
  group_by(species, treatment) %>% 
  count(name = "total")

species_ctmax_collections = acclim_data %>% 
  select(species, treatment, collection_date) %>% 
  distinct() %>% 
  group_by(species, treatment) %>% 
  count(name = "collections")

species_ctmax_mean = acclim_data %>% 
  group_by(species, treatment, collection_date) %>% 
  count() %>% 
  ungroup() %>% 
  group_by(species, treatment) %>% 
  summarise(mean_n = mean(n))

species_ctmax_total %>% 
  inner_join(species_ctmax_collections, by = c("species", "treatment")) %>% 
  inner_join(species_ctmax_mean, by = c("species", "treatment")) %>% 
  knitr::kable(digits = 2, caption = "Sample size information from the CTmax experiments. Shown here for each species and salt type are the total number of CTmax measurements, the number of collections (i.e. experimental replicates), and the average number of CTmax measurements per replicate")

```

### Body Sizes
```{r}

ggplot(acclim_data, aes(x = size)) + 
  facet_wrap(species~.) + 
  geom_histogram(fill = "grey", colour = "grey30",
                 binwidth = 0.05) + 
  labs(x = "Body Size (mm)") + 
  theme_matt_facets()

```


### Survival 
The salinity treatments exhibit the expected survival curves, with minimal acute effects, but strong effects emerging over five days.

```{r surv-curves, fig.width=8, fig.height=9}

surv_2025_data %>% 
  mutate(day = as.factor(day)) %>% 
  ggplot(aes(x = cl_conc, y = prop_surv, colour = day, group = day)) + 
  facet_grid(species ~ treatment) + 
  geom_hline(yintercept = 0.5, colour = "grey") + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "glm",
              method.args = list(family = "quasibinomial"), 
              se = F, 
              linewidth = 2.5) + 
  scale_y_continuous(breaks = c(0,1)) + 
  scale_colour_brewer(palette = "RdBu", direction = -1) + 
  labs(x = "[Chloride] (g / L)", 
       y = "Proportion Surviving") + 
  theme_bw(base_size = 24) + 
  theme(panel.grid = element_blank())
```

The survival curves were modeled as a logistic regression using a generalized linear mixed effects model with a binomial link function. The data was converted from group survival data (the number of individuals surviving in each dish) to individual binary survival data (1 = survived, 0 = dead), with the correct number of 1s and 0s based on the group survival data.

```{r}

surv_model_data = data.frame()
for(i in 1:dim(surv_2025_data)[1]){
  conditions = surv_2025_data[i,] %>% 
    dplyr::select(treatment, species, collection_date, collection_temp, cl_conc, time, day)
  
  if(conditions$species != "Skistodiaptomus"){
    initial = surv_2025_data$initial[i]
    surviving = surv_2025_data$surviving[i]
    dead = initial - surviving
    individual = c(1:initial)
    
    binary_data = data.frame("individual" = individual, 
                             "surv" = rep(c(1,0), times = c(surviving, dead)))
    
    day_data = bind_cols(conditions, binary_data) %>% 
      mutate(mort = if_else(surv == 1, 0, 1))
    
    surv_model_data = bind_rows(surv_model_data, day_data)
    
  }
  
}  

surv.model = glmmTMB::glmmTMB(surv ~ species * treatment * cl_conc * day +  (1 | collection_date), 
                              data = surv_model_data, 
                              family = binomial)


# 
# 
# performance::check_model(surv.model)
```

```{r}

summary(surv.model)

performance::r2_nakagawa(surv.model)
# Conditional = variance explained by both fixed and random effects

```

```{r}

car::Anova(surv.model, type = "III") %>% knitr::kable()
```

To clarify the patterns, we further analyzed the data using 1) model predictions of survival, 2) a Kaplan-Meier survival analysis, and 3) estimates of LC50 (the chloride concentration inducing 50% mortality).  

### Model Predictions
The generalized linear mixed model described above was used to predict survival data for the two species exposed to a range of chloride concentrations (0 - 3 g chloride), for 0-10 days. The predictions assume a collection temperature of 10°C and a collection date in mid-March.

The predicted survival values reinforce the main points from the raw survival data:

1.  L. minutus has a higher salt tolerance than L. sicilis.
2.  Both species are more sensitive to MgCl2 than NaCl.
3.  Survival decreases rapidly at higher chloride concentrations, but this varies across salts and species.

```{r}

pred_conc = rep(seq(from = 0, to = 3, by = 0.01), times = 4)
pred_treatments = rep(c("MgCl2", "NaCl"), times = length(pred_conc) / 2)
pred_species = rep(c("L. minutus", "L. sicilis"), each = length(pred_conc) / 2)

pred_data = data.frame(species = pred_species, 
                       treatment = pred_treatments, 
                       cl_conc = pred_conc, 
                       collection_temp = 10, 
                       day = rep(c(0:10), each = length(pred_species)), 
                       collection_date = "2026-03-16") %>% 
  mutate(surv = predict(surv.model, newdata = ., type="response"))

ggplot(pred_data, aes(x = cl_conc, y = surv, colour = day, group = day)) + 
  facet_grid(treatment~species) + 
  geom_line(linewidth = 2) + 
  scale_color_viridis_c(option = "G", direction = -1) + 
  theme_matt_facets()

```


### LC50 Values 
We subset the data for each species, treatment, and day. We fit a logistic regression to each subset, and then estimated the LC50 (chloride concentration where 50% mortality is observed). These LC50 values are shown below. Note: the NaCl LC50 value for L. sicilis on Day 1 is ~25 g Chloride, but has been cut off to highlight the patterns on later days.

```{r}

salt_cols = c("MgCl2" = "#761968", 
              "NaCl" = "#32ad3a")

# 2. Map model and dose.p to each nested data frame
lc50_values = surv_model_data %>%
  filter(day != 0) %>% 
  group_by(treatment, species, day) %>%
  nest() %>%
  mutate(
    model = map(data, ~glm(surv ~ cl_conc, # Fit logistic regression
                         data = .x, family = binomial)),
    lc50 = map(model, ~dose.p(.x, p = 0.5))) %>% # Extract dose.p for specific probability (e.g., 0.50 for LD50)
  dplyr::select(treatment, species, day, lc50) %>% 
  hoist(lc50, "p = 0.5:") %>%  
  dplyr::select(treatment:day, "lc50" = `p = 0.5:`)

lc50_values %>% 
ggplot(aes(x = day, y = lc50, colour = treatment)) + 
  facet_grid(treatment ~ species) + 
  geom_hline(yintercept = 0) + 
  geom_line(linewidth = 2) + 
  geom_point(size = 3) + 
  scale_color_manual(values = salt_cols) + 
  coord_cartesian(ylim = c(0,10)) + 
  theme_matt_facets() + 
  theme(legend.position = "none")

```


### Thermal Limit Effects 

Shown below are the average CTmax values for each chloride concentration (one mean per species, ion, and collection date combination). While salt acclimation has minimal effects on L. minutus, MgCl2 acclimation seems to have strong negative effects on L. sicilis. Note that the x-axis differs slightly between species.

```{r ctmax-salt-plot, fig.width=9, fig.height=4.5}

acclim_data %>% 
  group_by(species, 
           collection_date, 
           treatment, 
           cl_conc) %>% 
  summarise(mean_ctmax = mean(ctmax)) %>% 
  ggplot(aes(x = cl_conc, y = mean_ctmax, colour = treatment)) + 
  facet_wrap(.~species, scales = "free_x") + 
  geom_point(data = acclim_data, aes(x = cl_conc, y = ctmax, colour = treatment), 
             alpha = 0.2) + 
  geom_point(size = 3) + 
  geom_smooth(method = "lm", linewidth = 2) + 
  scale_colour_manual(values = salt_cols) + 
  scale_x_continuous(breaks = c(0,1,2), 
                     labels = c("0", "1", "2")) + 
  labs(y = "CTmax (°C)", 
       x = "Chloride (grams)") + 
  theme_matt_facets()
```

These same effects can be shown against the proportional mortality - this is meant to provide a standardized metric for "stress" of exposure to these different salts.

```{r ctmax-surv-plot, fig.width=9, fig.height=4.5}

surv_ctmax = acclim_data %>% 
  group_by(species, 
           collection_date, 
           treatment, 
           cl_conc) %>% 
  summarise(mean_ctmax = mean(ctmax)) %>% 
  mutate(collection_date = as_date(collection_date))

surv_surv = surv_2025_data %>% 
  group_by(treatment, species, collection_date) %>% 
  filter(time == max(time))

left_join(surv_ctmax, surv_surv) %>% 
  mutate(prop_mort = 1 - prop_surv) %>% 
  ggplot(aes(x = prop_mort, y = mean_ctmax, colour = treatment)) + 
  facet_wrap(.~species) + 
  geom_point(size = 3) +
  geom_smooth(method = "lm", linewidth = 2) + 
  scale_colour_manual(values = salt_cols) + 
  labs(y = "Mean CTmax (°C)", 
       x = "Proportion Mortality") + 
  scale_x_continuous(breaks = c(0,0.5,1), 
                     labels = c("0", "0.5", "1")) +
  theme_matt_facets()

```

We modeled these effects using a mixed effects model, with CTmax as a function of chloride concentration and salt compound (with interaction). Species was also included as a fixed effect, while collection date and tube position were included as random effects.

```{r}
acclim_data.model = lmer(data = acclim_data, 
                         ctmax ~ cl_conc * treatment * species + size + (1|collection_date))
```

```{r model-performance, fig.height=15, fig.width=11}
performance::check_model(acclim_data.model)

summary(acclim_data.model)

performance::r2_nakagawa(acclim_data.model)
```

Marginal trends were calculated based on this model, representing the effect of each salt type on CTmax (x axis shows the linear trend in CTmax per unit increase in chloride - a negative trend indicates that CTmax decreases as chloride concentrations increase). CTmax only seems to decrease substantially when L. sicilis is exposed to MgCl2.

```{r salt-contrasts, fig.width=9, fig.height=5}
car::Anova(acclim_data.model, type = "III")

emmeans::emtrends(acclim_data.model, var = "cl_conc", specs = "treatment", by = "species") %>% 
  data.frame() %>% 
  ggplot(aes(x = treatment, y = cl_conc.trend, colour = treatment)) + 
  facet_wrap(species~.) + 
  geom_hline(yintercept = 0, colour = "grey") + 
  geom_point(size = 5, position = position_dodge(width = 1)) + 
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
                width = 0.1, linewidth = 2,
                position = position_dodge(width = 1)) + 
  scale_colour_manual(values = salt_cols) + 
  labs(x = "Salt", 
       y = "CTmax vs. Chloride Trend \n(°C / g Chloride)") + 
  coord_flip() + 
  theme_matt_facets() + 
  theme(legend.position = "none")

```
