---
title: Effects of salinity acclimation on upper thermal limits of Lake Champlain diaptomid copepods
date: "`r Sys.Date()`"
output: 
  html_document:
          code_folding: hide
          code_download: true
          toc: true
          toc_float: true
  github_document:
          html_preview: false
          toc: true
          toc_depth: 3
---

```{r setup, include=T, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.align = "center",
  fig.path = "../Figures/report/",
  dev = c("png", "pdf"),
  message = FALSE,
  warning = FALSE,
  collapse = T
)

theme_matt = function(base_size = 18,
                      dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  ggpubr::theme_pubr(base_family="sans") %+replace% 
    theme(
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}

theme_matt_facets = function(base_size = 18,
                             dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  theme_bw(base_family="sans") %+replace% 
    theme(
      panel.grid = element_blank(),
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      strip.text.x = element_text(size = base_size),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}
```

## Survival Analyses

```{r}
ggplot(daily_prop_data, aes(x = treatment, y = prop_surv, colour = factor(exp_day))) + 
  facet_wrap(salt~.) + 
  geom_hline(yintercept = 0.5,
             colour = "grey", 
             linetype = "dashed") + 
  geom_point(size = 3) + 
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              se = FALSE,
              linewidth = 2) + 
  scale_colour_brewer(type = "seq", palette = 9) + 
  guides(colour = guide_legend(nrow = 1)) + 
  labs(x = "Salinity (mg/L)",
       y = "Proportion Surviving",
       colour = "Day") + 
  theme_matt_facets() + 
  theme(legend.position = "bottom")
```

```{r}
# surv_obj = Surv(surv_data$hour, surv_data$ind_surv)
# surv_fit = survfit2(Surv(hour, ind_surv) ~ treatment + salt, data = surv_data)
# 
# #summary(surv_fit_2)
# 
# ggsurvplot_facet(surv_fit, 
#                  data = surv_data,
#                  facet.by = "salt",
#                  conf.int=T, pval=F, risk.table=F, 
#                  conf.int.alpha = 0.1,
#                  size = 2,
#                  palette = "YlOrRd",
#                  legend.title="Salt Treatment")
```

```{r include = F}
cox.model = coxph(Surv(exp_day, ind_surv) ~ treatment + salt, data = surv_data)

cox.model

#ggforest(cox.model, data = surv_data)
```

### Skistodiaptomus oregonensis 

```{r, fig.width=10, fig.height=6}

oreg_data %>% 
  mutate(initial = hour_0) %>% 
  pivot_longer(cols = c("hour_0", "hour_18", "hour_42", "hour_62", "hour_86"), 
               names_to = "hour", 
               values_to = "individuals") %>% 
  mutate(prop_surv = individuals / initial) %>% 
  ggplot(aes(x = treatment, y = prop_surv, colour = sex)) +
  facet_wrap(hour~.) + 
  geom_point(position = position_jitter(height = 0.01, width = 200)) + 
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              se = FALSE,
              linewidth = 2) + 
  scale_colour_manual(values = c("female" = "lightcoral", 
                                 "male" = "lightblue3")) + 
  labs(x = "Salinity (mg/L)", 
       y = "Proportion Surviving") + 
  theme_matt_facets()

```

## CTmax Data
```{r include = F}
ctmax_data %>%  
  mutate("ID" = paste(experiment_date, "- Exp.", experiment)) %>% 
  ggplot(aes(x = treatment, y = ctmax, fill = ID)) +
  facet_wrap(salt~.) +
  geom_boxplot(width = 0.5,
               position = position_dodge(width = 0.7)) +
  geom_point(size = 3,
             position = position_dodge(width = 0.7)) + 
  labs(x = "Treatment", 
       y = "CTmax (°C)") + 
  guides(fill = guide_legend(override.aes = list(shape = NA))) + 
  theme_matt() + 
  theme(legend.position = "right")
```

Shown below are the sample sizes for the various salt-treatment combinations. Fewer experiments were run with Instant Ocean, so the sample sizes are smaller for this salt. 
```{r}
ctmax_filtered = ctmax_data %>% 
  filter(!(experiment == 1 & experiment_date == "1/20/24")) %>% 
  filter(!(experiment_date == "2/3/24")) %>% 
  mutate("ID" = paste(salt, treatment, sep = " - ")) %>% 
  mutate(group = case_when(
    treatment == "control" ~ "control",
    treatment == "salt" ~ salt,
    treatment == "salt" ~ salt
  ))

knitr::kable(ctmax_filtered %>% 
               group_by(salt, treatment) %>%  
               count())
```

A linear mixed effects model was used to examine variation in CTmax as a function of salt type (Instant Ocean and road salt) and treatment (control or salt-acclimated). We also examined the interaction between these factors to determine whether the effect of treatment depended on the type of salt used. Experiment date was included as a random effect to control for any differences between collections or experimental replicates. 

```{r}
salt.model = lmer(data = ctmax_filtered,
                  ctmax ~ treatment * salt + (1|experiment_date))

# salt.model = lm(data = ctmax_filtered,
#                 ctmax ~ treatment * salt)

knitr::kable(car::Anova(salt.model, type = "III"))
```

The mixed effects model indicates that there was a significant interaction between salt type and treatment. We examined this interaction by calculating the marginal treatment means by salt type. Shown below, there was no significant difference between treatment groups in the Instant Ocean experiments, but there was a substantial difference between treatment groups in the road salt experiments. 
```{r}
salt.means = emmeans::emmeans(salt.model,
                              pairwise ~ treatment | salt)

knitr::kable(salt.means$contrasts)
```

```{r plot-for-poster, fig.width = 8, fig.height=7}

treat_cols = c("salt" = "#95d5ce", "control" = "#028260")

ggplot(ctmax_filtered, aes(x = treatment, y = ctmax, fill = treatment)) +
  facet_wrap(salt~.) + 
  geom_boxplot(width = 0.5) +
  geom_point(size = 4, alpha = 0.5) + 
  scale_fill_manual(values = treat_cols) + 
  labs(x = "Treatment", 
       y = "CTmax (°C)") + 
  theme_matt_facets() + 
  theme(legend.position = "none",
        strip.background = element_rect(fill = "lightsalmon"),
        strip.text = element_text(margin = margin(0.7,0,0.7,0, "cm"),
                                  size = 20))

# ggplot(ctmax_filtered, aes(x = ctmax, fill = treatment)) + 
#   facet_wrap(salt~.) + 
#   geom_density(alpha = 0.5)
# 
# # Approach 1 - Simple t-test (comparison between two groups)
# t.test(data = filter(ctmax_filtered, salt == "RoadSalt"), ctmax ~ treatment)
```

```{r include = F}
ggplot(ctmax_filtered, aes(x = treatment, y = ctmax, fill = treatment)) +
  facet_wrap(salt~.) + 
  geom_violin(draw_quantiles = c(0.25,0.75)) + 
  geom_point(size = 4) + 
  labs(x = "Treatment", 
       y = "CTmax (°C)") + 
  theme_matt()
```

## Ion Specific Patterns Across Seasons

The salinity treatments exhibit the expected survival curves, with minimal acute effects, but strong effects emerging over five days. 

```{r surv-curves, fig.width=8, fig.height=9}

surv_2025_data %>% 
  mutate(day = as.factor(day)) %>% 
  ggplot(aes(x = cl_conc, y = prop_surv, colour = day, group = day)) + 
  facet_grid(species ~ treatment) + 
  geom_hline(yintercept = 0.5, colour = "grey") + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "glm",
              method.args = list(family = "quasibinomial"), 
              se = F, 
              linewidth = 2.5) + 
  scale_y_continuous(breaks = c(0,1)) + 
  scale_colour_brewer(palette = "RdBu", direction = -1) + 
  labs(x = "[Chloride] (mg / L)", 
       y = "Proportion Surviving") + 
  theme_bw(base_size = 24) + 
  theme(panel.grid = element_blank())
```

Shown below are the average CTmax values for each chloride concentration (one mean per species, ion, and collection date combination). While salt acclimation has minimal effects on L. minutus, MgCl2 acclimation seems to have strong negative effects on L. sicilis. Note that the x-axis differs slightly between species. 

```{r ctmax-salt-plot, fig.width=9, fig.height=4.5}

salt_cols = c("MgCl2" = "#761968", 
              "NaCl" = "#32ad3a")

acclim_data %>% 
  group_by(species, 
           collection_date, 
           treatment, 
           cl_conc) %>% 
  summarise(mean_ctmax = mean(ctmax)) %>% 
  ggplot(aes(x = cl_conc, y = mean_ctmax, colour = treatment)) + 
  facet_wrap(.~species, scales = "free_x") + 
  geom_point(data = acclim_data, aes(x = cl_conc, y = ctmax, colour = treatment), 
             alpha = 0.2) + 
  geom_point(size = 3) + 
  geom_smooth(method = "lm", linewidth = 2) + 
  scale_colour_manual(values = salt_cols) + 
  scale_x_continuous(breaks = c(0,1,2), 
                     labels = c("0", "1", "2")) + 
  labs(y = "CTmax (°C)", 
       x = "Chloride (grams)") + 
  theme_matt_facets()
```

These same effects can be shown against the proportional mortality - this is meant to provide a standardized metric for "stress" of exposure to these different salts. 

```{r ctmax-surv-plot, fig.width=9, fig.height=4.5}

surv_ctmax = acclim_data %>% 
  group_by(species, 
           collection_date, 
           treatment, 
           cl_conc) %>% 
  summarise(mean_ctmax = mean(ctmax)) %>% 
  mutate(collection_date = as_date(collection_date))

surv_surv = surv_2025_data %>% 
  group_by(treatment, species, collection_date) %>% 
  filter(time == max(time))

left_join(surv_ctmax, surv_surv) %>% 
  mutate(prop_mort = 1 - prop_surv) %>% 
  ggplot(aes(x = prop_mort, y = mean_ctmax, colour = treatment)) + 
  facet_wrap(.~species) + 
  geom_point(size = 3) +
  geom_smooth(method = "lm", linewidth = 2) + 
    scale_colour_manual(values = salt_cols) + 
  labs(y = "Mean CTmax (°C)", 
       x = "Proportion Mortality") + 
  scale_x_continuous(breaks = c(0,0.5,1), 
                     labels = c("0", "0.5", "1")) +
  theme_matt_facets()

```

We modeled these effects using a mixed effects model, with CTmax as a function of chloride concentration and salt compound (with interaction). Species was also included as a fixed effect, while collection date and tube position were included as random effects.  
```{r}
acclim_data.model = lme4::lmer(data = acclim_data, 
                               ctmax ~ cl_conc * treatment * species + (1|collection_date) + (1|tube))
```


```{r model-performance, fig.height=15, fig.width=11}
performance::check_model(acclim_data.model)
```

Marginal trends were calculated based on this model, representing the effect of each salt type on CTmax (x axis shows the linear trend in CTmax per unit increase in chloride - a negative trend indicates that CTmax decreases as chloride concentrations increase). CTmax only seems to decrease substantially when L. sicilis is exposed to MgCl2. 

```{r salt-contrasts, fig.width=9, fig.height=5}
#car::Anova(acclim_data.model, test = "F")

emmeans::emtrends(acclim_data.model, var = "cl_conc", specs = "treatment", by = "species") %>% 
  data.frame() %>% 
  ggplot(aes(x = treatment, y = cl_conc.trend, colour = treatment)) + 
  facet_wrap(species~.) + 
  geom_hline(yintercept = 0, colour = "grey") + 
  geom_point(size = 5, position = position_dodge(width = 1)) + 
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
                width = 0.1, linewidth = 2,
                position = position_dodge(width = 1)) + 
  scale_colour_manual(values = salt_cols) + 
  labs(x = "Salt", 
       y = "CTmax vs. Chloride Trend \n(°C / g Chloride)") + 
  coord_flip() + 
  theme_matt_facets() + 
  theme(legend.position = "none")

```

